---
title: "OS/PFS Survival Analysis — Cox Proportional Hazards Model"
subtitle: "TCGA Cohorts · Predefined Gene Signatures"
date: today
author: "Larkspur Survival Pipeline"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    toc-title: "Contents"
    number-sections: true
    code-fold: true
    code-summary: "Show code"
    theme: cosmo
    fig-width: 9
    fig-height: 6
    self-contained: true
    df-print: kable
execute:
  echo: true
  warning: false
  message: false
  cache: false
params:
  cohorts_yaml:   "metadata/cohorts.yaml"
  signatures_yaml: "metadata/signatures.yaml"
  data_dir:       "/media/kirovs/p1/cbio/datahub/public/"
---

```{r setup}
#| label: setup
#| include: false

# Load project utilities
source(here::here("R", "utils.R"))

library(kableExtra)
library(here)

knitr::opts_chunk$set(
  fig.align  = "center",
  out.width  = "95%"
)
```

## Introduction

This report performs **Overall Survival (OS)** and **Progression-Free Survival (PFS)** analysis
for predefined gene signatures across TCGA cohorts, using **Cox proportional hazards (Cox PH)** models.

For each combination of cohort × signature × endpoint, the analysis:

1. Loads clinical and mRNA expression data from the cBioPortal folder structure.
2. Computes a per-sample **signature score** (z-scored mean of constituent genes).
3. Dichotomises patients into **High** / **Low** groups at the cohort median score.
4. Fits a **Cox PH model** (unadjusted) and reports hazard ratio (HR), 95 % CI, and *p*-value.
5. Tests the **proportional hazards (PH) assumption** via Schoenfeld residuals.
6. Generates **Kaplan–Meier (KM) curves** and a **forest plot** across all analyses.

---

## Configuration

```{r load-config}
#| label: load-config

cat(getwd())
cat(params$cohorts_yaml)
cat(params$signatures_yaml)
cohort_cfg <- yaml::read_yaml(here(params$cohorts_yaml))
sig_cfg    <- yaml::read_yaml(here(params$signatures_yaml))

cat(sprintf("Cohorts loaded    : %d\n", length(cohort_cfg$cohorts)))
cat(sprintf("Signatures loaded : %d\n", length(sig_cfg$signatures)))
cat("\nCohorts:\n")
for (c in cohort_cfg$cohorts) cat(sprintf("  • %s (%s)\n", c$label, c$study_id))
cat("\nSignatures:\n")
for (s in sig_cfg$signatures) cat(sprintf("  • %s — genes: %s\n", s$label,
                                           paste(s$genes, collapse = ", ")))
```

---

## Data Loading

```{r load-data}
#| label: load-data

data_dir <- here(params$data_dir)

cohort_data <- map(cohort_cfg$cohorts, function(cfg) {
  clin_path <- file.path(data_dir, cfg$study_id, cfg$clinical_file)
  expr_path <- file.path(data_dir, cfg$study_id, cfg$expression_file)

  stopifnot(
    file.exists(clin_path),
    file.exists(expr_path)
  )

  list(
    config   = cfg,
    clinical = read_cbio_clinical(clin_path),
    expr     = read_cbio_expression(expr_path)
  )
})

names(cohort_data) <- map_chr(cohort_cfg$cohorts, "label")

# Quick data overview
overview_tbl <- imap_dfr(cohort_data, function(d, lbl) {
  tibble(
    Cohort         = lbl,
    Patients       = nrow(d$clinical),
    Genes          = nrow(d$expr),
    OS_events      = sum(d$clinical$OS_STATUS  == "1:DECEASED",    na.rm = TRUE),
    PFS_events     = sum(d$clinical$PFS_STATUS == "1:PROGRESSION",  na.rm = TRUE)
  )
})

kable(overview_tbl, caption = "Cohort overview") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

---

## Survival Analysis

```{r run-analysis}
#| label: run-analysis

all_results <- list()

for (cohort_name in names(cohort_data)) {
  d        <- cohort_data[[cohort_name]]
  cfg      <- d$config
  clinical <- d$clinical
  expr     <- d$expr

  all_results[[cohort_name]] <- list()

  for (endpoint in cfg$endpoints) {
    all_results[[cohort_name]][[endpoint]] <- list()

    for (sig in sig_cfg$signatures) {
      sig_name <- sig$name
      method   <- if (!is.null(sig$scoring)) sig$scoring else "mean"

      scores <- tryCatch(
        compute_signature_score(expr, sig$genes, method = method),
        error = function(e) {
          message(sprintf("[SKIP] %s / %s / %s: %s", cohort_name, endpoint, sig_name, e$message))
          NULL
        }
      )
      
      hist(scores,breaks=100,main=paste("Histogram of signature",sig_name, "calculated score in", cohort_name))
      
      if (is.null(scores)) next
    
      
      surv_data <- tryCatch(
        prepare_survival_data(clinical, scores, endpoint = endpoint),
        error = function(e) {
          message(sprintf("[SKIP] %s / %s / %s: %s", cohort_name, endpoint, sig_name, e$message))
          NULL
        }
      )
      if (is.null(surv_data) || nrow(surv_data) < 10) next

      cox_res <- tryCatch(
        run_cox_analysis(surv_data),
        error = function(e) {
          message(sprintf("[SKIP] %s / %s / %s Cox error: %s",
                          cohort_name, endpoint, sig_name, e$message))
          NULL
        }
      )
      if (is.null(cox_res)) next

      all_results[[cohort_name]][[endpoint]][[sig_name]] <- c(
        cox_res,
        list(
          surv_data     = surv_data,
          sig_label     = sig$label,
          sig_direction = if (!is.null(sig$direction)) sig$direction else NA_character_,
          endpoint      = endpoint
        )
      )
    }
  }
}

cat("Analysis complete. Results matrix:\n")
for (cohort in names(all_results)) {
  for (ep in names(all_results[[cohort]])) {
    sigs_done <- names(all_results[[cohort]][[ep]])
    cat(sprintf("  %s / %s : %s\n", cohort, ep, paste(sigs_done, collapse = ", ")))
  }
}
```

---

## Results per Cohort

```{r results-per-cohort}
#| label: results-per-cohort
#| results: asis

for (cohort_name in names(all_results)) {
  cat(sprintf("\n### %s\n\n", cohort_name))

  for (endpoint in names(all_results[[cohort_name]])) {
    ep_label <- if (endpoint == "OS") "Overall Survival (OS)" else
      "Progression-Free Survival (PFS)"
    cat(sprintf("\n#### %s\n\n", ep_label))

    for (sig_name in names(all_results[[cohort_name]][[endpoint]])) {
      res       <- all_results[[cohort_name]][[endpoint]][[sig_name]]
      sig_label <- res$sig_label
      surv_data <- res$surv_data

      cat(sprintf("\n##### %s\n\n", sig_label))

      # --- Cox summary table ---
      cat("**Cox PH model — High vs Low signature score:**\n\n")
      tbl <- res$summary_tbl %>%
        mutate(across(c(hr, ci_lower, ci_upper), ~round(., 3)),
               p_value = fmt_pval(p_value)) %>%
        rename(Term = term, HR = hr, `CI lower` = ci_lower,
               `CI upper` = ci_upper, `P-value` = p_value)
      print(kable(tbl, format = "html") %>%
              kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                            full_width = FALSE))
      cat("\n")

      # --- PH assumption test ---
      if (!is.null(res$ph_test)) {
        ph_df <- as.data.frame(res$ph_test$table)
        ph_df <- ph_df %>%
          rownames_to_column("term") %>%
          mutate(across(where(is.numeric), ~round(., 4)))
        cat("**Proportional hazards test (Schoenfeld residuals):**\n\n")
        ph_p <- ph_df$p[ph_df$term == "GLOBAL"]
        if (!is.na(ph_p) && ph_p < 0.05) {
          cat("::: {.callout-warning}\n")
          cat(sprintf("PH assumption may be violated (global p = %.3f).", ph_p))
          cat(" Consider time-varying covariate or stratified model.\n")
          cat(":::\n\n")
        } else {
          cat(sprintf("Global Schoenfeld test p = %.3f — PH assumption not violated.\n\n",
                      if (is.na(ph_p)) NA_real_ else ph_p))
        }
      }

      # --- KM plot ---
      km <- tryCatch(
        plot_km_curves(
          surv_data,
          title    = sprintf("%s | %s | %s", cohort_name, endpoint, sig_label),
          endpoint = endpoint
        ),
        error = function(e) NULL
      )
      if (!is.null(km)) {
        print(km)
        cat("\n")
      }
    }
  }
}
```

---

## Summary

### Results Table

```{r summary-table}
#| label: summary-table

summary_tbl <- build_summary_table(all_results) %>%
  rename(
    `CI Lower`    = CI_lower,
    `CI Upper`    = CI_upper,
    `P-value`     = P_value,
    `Direction OK` = Direction_OK
  ) %>%
  mutate(
    `Direction OK` = dplyr::case_when(
      is.na(`Direction OK`)    ~ "—",
      isTRUE(`Direction OK`)   ~ "\u2714",  # checkmark
      TRUE                     ~ "\u2718"   # cross
    )
  )

kable(
  summary_tbl,
  caption = "Cox PH results — High vs Low signature score across all cohorts and endpoints.\nDirection OK: whether the observed HR matches the expected direction declared in signatures.yaml."
) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  column_spec(4, bold = TRUE) %>%
  row_spec(which(as.numeric(gsub("<", "", summary_tbl$`P-value`)) < 0.05 |
                   summary_tbl$`P-value` == "<0.001"), background = "#FFF3CD")
```

### Forest Plots

```{r forest-plots}
#| label: forest-plots
#| fig-height: 5

for (cohort_name in names(all_results)) {
  for (endpoint in names(all_results[[cohort_name]])) {
    ep_res <- all_results[[cohort_name]][[endpoint]]

    if (length(ep_res) == 0) next

    # Build named list for forest plot
    named_res <- set_names(ep_res, map_chr(ep_res, "sig_label"))

    p <- plot_forest(named_res)
    p <- p + labs(
      title    = sprintf("Forest Plot — %s | %s", cohort_name, endpoint),
      subtitle = "High vs Low signature score · Cox PH (unadjusted)"
    )
    print(p)
    cat("\n")
  }
}
```

---

## Methods

### Statistical Methods

**Signature scoring:** Gene expression values (log2 RSEM) are z-scored per gene
across all samples within a cohort. The per-sample signature score is the arithmetic
mean of z-scored values across all signature genes.

**Patient stratification:** Patients are divided into *High* and *Low* groups at the
cohort-specific **median** signature score.

**Cox proportional hazards model:** An unadjusted Cox PH model is fit with
`score_group` (High vs Low) as the predictor:

$$h(t) = h_0(t) \cdot \exp(\beta \cdot \text{score\_group})$$

Hazard ratios (HR), 95 % confidence intervals, and two-sided Wald test *p*-values
are reported.

**PH assumption:** The proportional hazards assumption is tested using scaled
Schoenfeld residuals (`cox.zph`). A global *p* < 0.05 indicates a potential violation.

**Kaplan–Meier curves:** KM estimates are plotted per group with 95 % pointwise
confidence bands. Log-rank test *p*-values are overlaid.

### Software

```{r session-info}
#| label: session-info
#| code-fold: false

si <- sessionInfo()
cat(sprintf("R version : %s\n", si$R.version$version.string))
pkgs <- c("survival", "survminer", "dplyr", "ggplot2", "yaml", "kableExtra")
for (pkg in pkgs) {
  v <- tryCatch(as.character(packageVersion(pkg)), error = function(e) "not installed")
  cat(sprintf("  %-15s %s\n", pkg, v))
}
```

---

*Report generated with [Quarto](https://quarto.org/) on `r format(Sys.time(), "%Y-%m-%d %H:%M %Z")`.*
